---
title: "Assignment 4"
subtitle: "Due at 11:59pm on November 7."
format: pdf
editor: visual
---

This is an individual assignment. Turn in this assignment as an HTML or PDF file to ELMS. Make sure to include the R Markdown or Quarto file that was used to generate it. Include the GitHub link for the repository containing these files.

-   **Link to the repository:** <https://github.com/annechang11/surv727-hw4>

```{r}
#| include: false 
# options(httr_oauth_cache = FALSE)
# options(gargle_oauth_cache = FALSE)
library(tidyverse)
library(DBI)
library(dbplyr)
library(bigrquery)
```

```{r, echo=FALSE}
# require(devtools)
# install_version("dbplyr", version = "2.3.4", repos = "http://cran.us.r-project.org")
```

In this notebook we will use Google BigQuery, "Google's fully managed, petabyte scale, low cost analytics data warehouse". Some instruction on how to connect to Google BigQuery can be found here: <https://db.rstudio.com/databases/big-query/>.

You will need to set up a Google account with a project to be able to use this service. We will be using a public dataset that comes with 1 TB/mo of free processing on Google BigQuery. As long as you do not repeat the work in this notebook constantly, you should be fine with just the free tier.

Go to <https://console.cloud.google.com> and make sure you are logged in a non-university Google account. **This may not work on a university G Suite account because of restrictions on those accounts.** Create a new project by navigating to the dropdown menu at the top (it might say "Select a project") and selecting "New Project" in the window that pops up. Name it something useful.

After you have initialized a project, paste your project ID into the following chunk.

```{r}
project <- "surv-727-test-403119"
```

We will connect to a public database, the Chicago crime database, which has data on crime in Chicago.

```{r}
con <- dbConnect(
  bigrquery::bigquery(),
  project = "bigquery-public-data",
  dataset = "chicago_crime",
  billing = project
)
con
```

We can look at the available tables in this database using `dbListTables`.

**Note**: When you run this code, you will be sent to a browser and have to give Google permissions to Tidyverse API Packages. **Make sure you select all to give access or else your code will not run.**

```{r}
dbListTables(con)
```

Information on the \`crime\` table can be found here:

<https://console.cloud.google.com/marketplace/product/city-of-chicago-public-data/chicago-crime?project=surv-727-test-403119>

Write a first query that counts the number of rows of the \`crime\` table in the year 2016. Use code chunks with {sql connection = con} in order to write SQL code within the document.

```{sql connection = con}
SELECT COUNT (*)
FROM crime
WHERE year = 2016;
```

-   There are 269839 (or 269840) rows in the year 2016 in the "crime" table.

Next, count the number of arrests grouped by `primary_type` in 2016. Note that is a somewhat similar task as above, with some adjustments on which rows should be considered. Sort the results, i.e. list the number of arrests in a descending order.

```{sql connection = con}
SELECT primary_type, count(*)
FROM crime
WHERE arrest = TRUE AND year = 2016
GROUP BY primary_type
ORDER BY count(*) DESC;

```

-   Please see the number of arrests in 2016 by primary type above.

We can also use the `date` for grouping. Count the number of arrests grouped by hour of the day in 2016. You can extract the latter information from `date` via `EXTRACT(HOUR FROM date)`. Which time of the day is associated with the most arrests?

```{sql connection = con}
SELECT EXTRACT(HOUR FROM date) AS hour, count(*)
FROM crime 
WHERE arrest = TRUE AND year = 2016
GROUP BY hour
ORDER BY count(*) DESC;
```

-   From the list above, we can see that hour 10 has the highest number of arrests (5306), following by hour 11 (5200). Note that we can't distinguish AM and PM from the data though.

Focus only on `HOMICIDE` and count the number of arrests for this incident type, grouped by year. List the results in descending order.

```{sql connection = con}
SELECT year, count(*)
FROM crime
WHERE arrest = TRUE AND primary_type = 'HOMICIDE'
GROUP BY year
ORDER BY count(*) DESC;

```

-   The list above shows the numbers of "Homicide" arrests in each year. Year 2001 has the highest number of arrests for homicide from year 2001 to 2022.

Find out which districts have the highest numbers of arrests in 2015 and 2016. That is, count the number of arrests in 2015 and 2016, grouped by year and district. List the results in descending order.

```{sql connection = con}
SELECT year, district, count(*)
FROM crime
WHERE arrest = TRUE AND year IN (2015, 2016)
GROUP BY year, district
ORDER BY count(*) DESC;
```

-   District 11 has the highest number of arrests in both 2015 and 2016.

Lets switch to writing queries from within R via the `DBI` package. Create a query object that counts the number of arrests grouped by `primary_type` of district 11 in year 2016. The results should be displayed in descending order.

Execute the query.

```{sql connection = con}
SELECT primary_type, count(*)
FROM crime
WHERE arrest = TRUE AND year = 2016 AND district = 11
GROUP BY primary_type
ORDER BY count(*) DESC;
  
```

```{r}
# store the above query in a R object

query <-
  "SELECT primary_type, count(*)
  FROM crime
  WHERE arrest = TRUE AND year = 2016 AND district = 11
  GROUP BY primary_type
  ORDER BY count(*) DESC;"

# run the query
dbGetQuery(con, query)
```

```{r}
subtable <- dbGetQuery(con, query)
str(subtable) # a tibble of 27*2
```

Try to write the very same query, now using the `dbplyr` package. For this, you need to first map the `crime` table to a tibble object in R.

```{r}
crime <- tbl(con, 'crime')
```

Again, count the number of arrests grouped by `primary_type` of district 11 in year 2016, now using `dplyr` syntax.

```{r}
crime %>%
  filter(year == 2016 & district == 11 & arrest == TRUE) %>%
  group_by(primary_type) %>%
  summarise(total = n())

# matches the sql result
```

Count the number of arrests grouped by `primary_type` and `year`, still only for district 11. Arrange the result by `year`.

```{r}
crime %>%
  filter(district == 11 & arrest == TRUE) %>%
  group_by(year, primary_type) %>%
  summarise(total = n()) %>%
  arrange(year, primary_type)

# note: I arranged the result by year from 2001 to 2022
# and then by primary_type alphebetically
```

Assign the results of the query above to a local R object.

```{r}
sql_data <-
  crime %>%
  filter(district == 11 & arrest == TRUE) %>%
  group_by(year, primary_type) %>%
  summarise(total = n()) %>%
  arrange(year, primary_type) %>%
  collect()
```

Confirm that you pulled the data to the local environment by displaying the first ten rows of the saved data set.

```{r}
head(sql_data, 10)
```

Close the connection.

```{r}
dbDisconnect(con)
```
